A Simple Type Theory for Category Theory

### 2022-04-11 paper/paper.tex
comment out the internal category theory stuff

  We show that the type theory is sound for models called
  \emph{virtual equipments}, a kind of double category which is known
  to model not just ordinary categories but enriched, internal and
  generalized multi-categories as well. This means that while proofs
  and constructions in our theory read like typical category-theoretic
  constructions, the proofs automatically apply to these generalized
  categories. Our type theory simplifies arguments over the algebraic
  description of virtual equipments by modeling the algebraic
  operation of restriction of profunctors along functors as instead
  the strict syntactic substitution.
\end{abstract}

@@

At this level of generality, the proofs can be difficult to
understand, as they often involve some combination of the original
category theoretic proof with some additional side-conditions that
everythinng is in fact appropriately step-indexed, continuous etc.
%
The solution to taming this complexity is to prove these general
theorems at a higher level of generality, for instance, proving them
in some abstract 2-category of enriched/internal categories, functors
and transformations.
%
For proofs in generalized category theory specifically, the notion of
a \emph{virtual equipment} has been shown to be occupy an algebraic
sweet-spot of being complex enough to express most basic constructions
(functors, profunctors, natural transformations, adjunctions, kan
extensions, etc) but also general enough to encompass internal,
enriched and ``generalized multi-categories'', which include ordinary
flavors of multi-categories like the cartesian multi-categories that
model sequents in natural deduction, but even structures such as
topological spaces.
%
However, there are two difficulties with this approach.
%
First, the abstract arguments can be difficult to understand, and can
seem quite hard to connect to the original argument for ordinary
category theory.
%
Second, the algebraic structure of a virtual equipment is quite
unwieldy to manipulate. The equivalent of a 2-cell is a kind of
``multi-2-cell'' with many inputs and a single output, in
diagrammatic form these are rendered as

@@

We solve both of these issues by developing an alternative \emph{type
theoretic syntax} for the constructions in a virtual equipment.
%
In this syntax the ``multi-2-cell'' above instead becomes a natural
transformation judgment
\[ \Phi \vdash t : R[F(x);G(y)] \]
We write natural transformations in a natural deduction style: a
transformation is written as a ``parameterized'' element of the output
profunctor. Furthermore the restriction operators for vertical arrows
into horizontal arrows is naturally modeled by a syntactic
substitution.
%
This interpretation of restriction as substitution both matches the
semantic intuition for the restriction operation as well as greatly
simplifies reasoning about composition of transformations.

@@

@@

%% \subsection{Products}

%% We add in parallel two different notions of product: products of sorts
%% and products of spans, which have the universal property in their
%% respective categories.

%% \begin{figure}
%%   \begin{mathpar}
%%     \inferrule{}{\Gamma \vdash 1 : \Sort}\and
%%     \inferrule*{}{\Gamma\pipe\alpha:C \vdash () : 1}\and
%%     \inferrule*{\Gamma\pipe\alpha:C \vdash A : 1}{A = ()}
%%     \inferrule{}{\Gamma\pipe \Xi \vdash 1 \isaSet}\and
%%     \inferrule{}{\Gamma\pipe \Phi \vdash () :  1}\and
%%     \inferrule{\Gamma\pipe \Phi \vdash M :  1}{M = ()}\\\\

%%     \inferrule{\Gamma \vdash C : \Sort \and \Gamma \vdash D : \Sort}
%%               {\Gamma \vdash C \times D : \Sort}\and

%%     \inferrule{\forall i \in \{1,2\}.~ \Gamma \pipe \alpha:C \vdash A_i : D_i}{\Gamma \pipe \alpha:C \vdash (A_1,A_2) : D_1 \times D_2}

%%     \inferrule{\Gamma \pipe \alpha:C \vdash A : D_1 \times D_2}{\Gamma \pipe \alpha:C \vdash \pi_i A : D_i}

%%     \pi_i(A_1,A_2) = A_i\\
%%     \inferrule{\Gamma\pipe \alpha:C \vdash A :  D_1\times D_2}{A = (\pi_1 A,\pi_2 A)}\\\\

%%     \inferrule{\Gamma\pipe\Xi \vdash R \isaSet \and \Gamma\pipe\Xi \vdash S\isaSet}
%%               {\Gamma\pipe\Xi \vdash R \times S\isaSet}\and

%%     \inferrule{\forall i \in \{1,2\}.~ \Gamma \pipe\Phi \vdash M_i : R_i}{\Gamma \pipe\Phi \vdash (M_1,M_2) : R_1 \times R_2}

%%     \inferrule{\Gamma \pipe\Phi\vdash M : R_1 \times R_2}{\Gamma \pipe\Phi \vdash \pi_i M : R_i}

%%     \pi_i(M_1,M_2) = M_i\\
%%     \inferrule{\Gamma\pipe\Phi \vdash M :  R_1\times R_2}{M = (\pi_1 M,\pi_2 M)}
%%   \end{mathpar}
%%   \caption{Products}
%%   \label{fig:products}
%% \end{figure}

%% One additional connective we might desire is a ``parallel'' product of
%% sets $R \times' S : \harr {C_1 \times C_2} {D_1 \times D_2}$ where $R
%% : \harr {C_1} {D_1}$ and $R : \harr {C_2} {D_2}$. We don't need to
%% define this as a primitive though because it is expressible in terms
%% of the two kinds of products we have above:

%% \[
%% \inferrule
%% {\Gamma \pipe \alpha : C_i; \beta : D_i \vdash R_i \isaSet}
%% {\Gamma \pipe \alpha : C_1\times C_2; \beta : D_1 \times D_2
%%   R_1[\pi_1\alpha/\alpha;\pi_1\beta/\beta] \times R_2[\pi_2\alpha/\alpha;\pi_2\beta/\beta]\isaSet}
%% \]

%% The empty products give us access now to several traditional notions
%% of category theory that we couldn't before define, such as objects and
%% presheaves of either variance:

%% $$\textrm{Object}\,C = \varr 1 C$$
%% $$\textrm{Presheaf}\,C = \harr 1 C$$
%% $$\textrm{coPresheaf}\,C = \harr C 1$$

%% However, we find that in practice we don't have much use for these
%% concepts as most theorems about them easily generalize to more general
%% theorems about arbitrary functors or profunctors.

@@

\begin{enumerate}
\item A functor $\textrm{wp} : \exists (\alpha:\cat B; \beta:\cat E). \punitinXfromYtoZ {\cat B} \alpha {p\beta}$
\item In the slice category: $p(\textrm{wp}(k)) = \pi_-k$
\item With a universal morphism $\epsilon : \pforallXfromYtoZ {k} \punitinXfromYtoZ {} {(\pi_-k,k,\punitrefl{}{\pi_-k})} {k}$
\item In the slice category: $\pi_k \epsilon = \id$ (how to express this nicely in SCT?)
\end{enumerate}

@@

\section{Internal Category Theory}

A downside of working synthetically is that since everything we define
must be functorial/natural, we cannot prove any theorems about
functions that are not assumed to be functorial or transformations
that are not assumed ot be natural.
%
These kinds of constructions are surprisingly common in basic category theory.
%
For instance, when defining a left adjoint $F$ to a given right
adjoint functor $U$, the functoriality of $F$ is often not assumed,
rather we are given that each $F a$ has a certain universal mapping
property and this can then be used to show the action of $F$ on
objects extends to an action on morphisms.
%
This definition is quite important for applications of category theory
to type theory: in type theory when a type constructor is defined it
is given introduction, elimination and $\beta\eta$ rules, but no
primitive syntax is given for a functorial action.
%
The fact that functoriality is determined from the universal mapping
property of the type is then a key step in proving the type theory is
sound and complete with respect to models which are expected to have
certain functors on them.

As another example, a common lemma states that an \emph{possibly
unnatural} transformation is in fact natural if it is the inverse of a
natural transformation.
%
Neither of these examples can be stated as synthetic theorems, as it
is impossible to quantify over ``possibly unnatural transformations''
or ``functions on objects''.
%
Inside the calculus, we cannot ``see'' that categories have an
underlying set of objects.
%
Fortunately there is a way to use our calculus in a different manner
to study these constructions: by defining categories \emph{internal
to} SCT.
%
This reproduces classical constructions, since there is a model where
a category $\cat C$ is interpreted simply as a set, functors as
functions and profunctors $\harr {\cat C}{\cat D}$ as \emph{spans},
i.e. sets $R$ with (source and target) projections $\cat C \leftarrow
R \rightarrow \cat D$.
%
The proofs read like the classical proofs (with some care additional
attention on the ordering of variable usage), but by virtue of being
syntactic automatically apply to any notion of category internal to a
virtual equipment, which includes classical internal categories,
enriched categories and a broad range of generalized multicategory.

We note that a synthetic approach to model this situation may be
possible were we to extend our calculus with a \emph{modality} $|\cat
C|$ that takes a category to the discrete category on the set of
objects. We discuss this in \cref{sec:discussion}.

Next, we introduce the notion of category internal to our syntactic
virtual equipment and the \emph{monoids and modules} translation that
shows that, with the inclusion of hom sets, our syntax is already
suitable for doing syntactic category theory.

The benefits of our syntax become clear when we define internal
categories, functors, profunctors and transformations: the definitions
read exactly as the classical definitions.
\begin{definition}[Internal Category]
  An \emph{(internal) category} consists of\footnote{the following is
    easily interpreted as a $\Sigma$ type}:
  \begin{enumerate}
  \item A sort $C_0 \isaCat$
  \item A span $C_1 : \harr{C_0}{ C_0}$
  \item Identity arrows
    $C_r : \pendallXdotY {\alpha :  C_0} {\harrapp{C_1} \alpha \alpha}$
  \item Composition of arrows
    \[ C_t : \pendallXdotY \alpha {\homrallXYtoZ \beta {\harrapp{C_1} \alpha \beta} {\homrallXYtoZ \gamma {\harrapp{C_1} \beta \gamma} {\harrapp{C_1} \alpha \gamma}}}\]

    %% \[ C_t^\triangleright : \forall \alpha :  C_0. \forall \beta. ({C_1} \beta \alpha) \triangleright \forall \gamma. ({C_1} \gamma \beta) \triangleright {C_1} \gamma \alpha \]

    %% or

    %% \[ C_t^\triangleleft : \forall \alpha. \forall \beta. (\forall \gamma. {C_1} \alpha \gamma \triangleleft {C_1} \beta \gamma) \triangleleft {C_1} \alpha \beta
    %% \]

    %% or using tensors and covariant homs

    %% \[ C_t^\odot : \forall \alpha. \forall \gamma. (\exists \beta. {C_1}\alpha\beta\odot {C_1}\beta\gamma)\triangleright {C_1}\alpha \gamma \]

    %% or using tensors and contravariant homs

    %% \[ C_t^\odot : \forall \gamma. \forall \alpha. {C_1}\alpha\gamma \triangleleft (\exists \beta. {C_1}\alpha\beta\odot {C_1}\beta\gamma)\]
  \item Satisfying right unitality, i.e., for any $\Phi \vdash f :
    \harrapp {C_1} A B$, we'd like $f; r(B) = f$. We can formalize this
    using an equality of two transformations of type $\pendallXdotY
    \alpha {\homrallXYtoZ \beta {\harrapp{C_1} \alpha \beta} {\harrapp{C_1} \alpha \beta}}$

    \[ \pendlambdaXdotY \alpha {\homrlambdaXatYdotZ f \beta {\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t} \alpha} {f} \beta} {\pendappXtoY {C_r} \beta} \beta}}
    = \pendlambdaXdotY \alpha {\homrlambdaXatYdotZ f \beta f} \]

    These kinds of equalities between 2-cells are very common in
    internal category theory, so to save space we will suppress the
    sequence of $\lambda$s that start such equalities. So the above can simply be written as

    \[ { {\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t} \alpha} {f} \beta} {\pendappXtoY {C_r} \beta} \beta}} = { f}\]

  \item and left unitality (an equality of the same type as right unitality):

    \[ { {\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t} \alpha} {\pendappXtoY {C_r} \alpha} \alpha} f \beta}} = { f}\]
  \item and associativity, which is an equality at type
    \[ \pendallXdotY \alpha {\homrallXYtoZ \beta {\harrapp{C_1} \alpha \beta} {\homrallXYtoZ \gamma {\harrapp{C_1} \beta \gamma} {\homrallXYtoZ \delta {\harrapp{C_1} \gamma \delta} {\harrapp{C_1} \alpha \delta}}}}\]

    \[ \homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t}\alpha} f \beta} {(\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t}\beta} g \gamma} h \delta)} \delta
    =
    \homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t}\alpha} {(\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t}\alpha} f \beta} g \gamma)} \gamma} h \delta \]
  \end{enumerate}
\end{definition}

We can define an internal functor to be a pair of an object function
and a function on arrows that preserves identity and 
\begin{definition}[Internal Functor]
  A functor $F$ between internal categories $C$ and $D$ consists of
  \begin{enumerate}
  \item a map of objects $F_0 : \varr {C_0} {D_0}$
  \item a map of arrows
    $F_1 : \pendallXdotY {\alpha}
    {\homrallXYtoZ \beta {\harrapp{C_1} \alpha \beta} {\harrapp{D_1} {F_0 \alpha}{F_0 \beta}}}$
  \item That preserves identity. This is given by equality between the
    following terms of type $\forall \alpha:{{C_0}}. D_1 {(F_0 \alpha)}{(F_0\alpha)}$:
    \[ \homrappXtoYatZ {\pendappXtoY {F_1}\alpha} {\pendappXtoY {C_r} \alpha} \alpha = \pendappXtoY {D_r}{F_0\alpha} \]
  \item And preserves composition.
    This is given by an equality between the following terms of type

    \[ \pendallXdotY \alpha
    \homrallXYtoZ {\beta} {\harrapp{C_1} \alpha \beta} {\homrallXYtoZ \gamma {\harrapp{C_1} \beta \gamma} {\harrapp{D_1} {F_0 \alpha} {F_0 \gamma}}}\]

    First, we can compose in $C$, and then map using $F$:
    \[ \homrappXtoYatZ {\pendappXtoY{F_1} \alpha} {(\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t} \alpha} f \beta} g \gamma)}\gamma\]

    or we can map the morphisms with $F$, and then compose them in $D$:
    \[ \homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {D_t} {(F_0 \alpha)}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} f \beta)} {(F_0 \beta)}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \beta} g \gamma)} {{F_0} \gamma}\]
  \end{enumerate}
\end{definition}

We can immediately define identity and composition of functors.

\begin{definition}
  For any $C : \textrm{Cat}$, we can define an identity functor
  $\textrm{Id}_C : \textrm{Functor} C C$ by
  \begin{enumerate}
  \item $\textrm{Id}_0 = \lambda \alpha:C_0. \alpha$
  \item $\textrm{Id}_1 = \pendlambdaXdotY \alpha
    {\homrlambdaXatYdotZ f {\alpha'} f}$
  \item Preservation of identity requires proving
    \[
    \lambda \alpha. \pendappXtoY {C_r}{\textrm{Id}_0\alpha} \]
    is equal to
    \[ \lambda \alpha.
    \homrappXtoYatZ {\pendappXtoY {\textrm{Id}_1} \alpha} {(\pendappXtoY {C_r}\alpha)} \alpha\]
    Both sides simply $\beta$ reduce to $ \lambda \alpha. \pendappXtoY {C_r} \alpha$
    and the result follows by reflexivity.
  \item Preservation of composition requires proving (under a $\lambda \alpha. \homrlambdaXatYdotZ f \beta \homrlambdaXatYdotZ g \gamma$)
    \[
    \homrappXtoYatZ {\pendappXtoY {{\textrm{Id}}_1} \alpha} {(\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t} \alpha} f \beta} g \gamma)}\gamma
    = \homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t} {{\textrm{Id}}_0 \alpha}} {(\homrappXtoYatZ {\pendappXtoY {{\textrm{Id}}_1} \alpha} f \beta)} {{\textrm{Id}}_0 \beta}} {(\homrappXtoYatZ {\pendappXtoY{{\textrm{Id}}_1} \beta} g \gamma)} {{\textrm{Id}}_0 \gamma}\]
    Both sides $\beta$ reduce to

    \[ \homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t} \alpha} f \beta} g \gamma\]
    so the result holds by reflexivity.
  \end{enumerate}
\end{definition}

\begin{definition}
  Given categories $C,D,E$ and functors $F : \textrm{Functor} C D$ and
  $G : \textrm{Functor} D E$, we define the composition $G \circ F :
  \textrm{Functor} C E$ as
  \begin{enumerate}
  \item $(G \circ F)_0 = \lambda \alpha. {G_0 ({F_0} \alpha)}$
  \item $(G \circ F)_1 =
    \pendlambdaXdotY \alpha {
    \homrlambdaXatYdotZ f \beta
    {\homrappXtoYatZ {\pendappXtoY {G_1} {F_0 \alpha}} {\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} f \beta} {F_0 \beta}}}$
  \item Preservation of identity requires proving
    \[ \lambda \alpha. \pendappXtoY {E_r} {(G \circ F)_0\alpha} \]
    is equal to
    \[ \lambda \alpha.
    \homrappXtoYatZ {\pendappXtoY {(G \circ F)_1} \alpha} {\pendappXtoY {C_r}\alpha} \alpha \]

    Expanding definitions and $\beta$ reducing we need to show

    \[ \pendlambdaXdotY\alpha {\pendappXtoY {E_r}{G_0 (F_0 \alpha)}} \]
    is equal to
    \[ \pendlambdaXdotY \alpha {\homrappXtoYatZ {\pendappXtoY {G_1} {F_0 \alpha}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} {\pendappXtoY {C_r} \alpha} \alpha)} {F_0 \alpha}}\]

    Proceeding from right to left applying transitivity:
    \begin{align*}
      &\pendlambdaXdotY \alpha {\homrappXtoYatZ {\pendappXtoY {G_1} {F_0 \alpha}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} {\pendappXtoY {C_r} \alpha} \alpha)} {F_0 \alpha}}\\
      &=\pendlambdaXdotY \alpha {\homrappXtoYatZ {\pendappXtoY {G_1} {F_0 \alpha}} {(\pendappXtoY{(\pendlambdaXdotY \alpha {\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} {\pendappXtoY {C_r} \alpha} \alpha})}\alpha)}} {F_0 \alpha}\\
      &=\pendlambdaXdotY \alpha {\homrappXtoYatZ {\pendappXtoY {G_1} {F_0 \alpha}} {(\pendappXtoY{(\pendlambdaXdotY \alpha {{\pendappXtoY {D_r} {F_0 \alpha}}})}\alpha)}} {F_0 \alpha}\tag{*}\\
      &=\pendlambdaXdotY \alpha {\homrappXtoYatZ {\pendappXtoY {G_1} {F_0 \alpha}} {\pendappXtoY {D_r} {F_0 \alpha}}} {F_0 \alpha}\\
      &= \pendlambdaXdotY \alpha
      \pendappXtoY {(\pendlambdaXdotY \beta {{\homrappXtoYatZ {\pendappXtoY {G_1} \beta} {\pendappXtoY {D_r} \beta} \beta}})} {F_0 \alpha}\\
      &= \pendlambdaXdotY \alpha
      \pendappXtoY {(\pendlambdaXdotY \beta \pendappXtoY {E_r} {G_0 \beta})} {F_0 \alpha}\tag{*}\\
      &= \pendlambdaXdotY \alpha \pendappXtoY {E_r} {G_0 (F_0 \beta)}\\
    \end{align*}

    Where the two marked lines follow from congruence using the
    assumptions about $F,G$.

  \item Preservation of Composition.

    \[ \homrappXtoYatZ {\pendappXtoY{(G \circ F)_1} \alpha} {(\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t} \alpha} f \beta} g \gamma)}\gamma\]
    equals

    \[ \homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {E_t} {(G \circ F)_0 \alpha}} {(\homrappXtoYatZ {\pendappXtoY {(G\circ F)_1} \alpha} f \beta)} {(G\circ F)_0 \beta}} {(\homrappXtoYatZ {\pendappXtoY {(G\circ F)_1} \beta} g \gamma)} {{F_0} \gamma}\]

    Reducing, we need to show
    \[ \homrappXtoYatZ {\pendappXtoY{G_1} {F_0 \alpha}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} {(\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t}\alpha} f {\beta}} g {\gamma})} \gamma)} {F_0 \gamma} \]

    is equal to
    \[ \homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {E_t} {G_0(F_0\alpha)}}
      {({\homrappXtoYatZ {\pendappXtoY {G_1} {F_0 \alpha}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} f \beta)} {F_0 \beta}})}
      {G_0(F_0\beta)}}
       {({\homrappXtoYatZ {\pendappXtoY {G_1} {F_0 \beta}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \beta} g \gamma)} {F_0 \gamma}})}
       {{F_0} \gamma}\]

    Going from left to right:
    \begin{align*}
      &\homrappXtoYatZ {\pendappXtoY{G_1} {F_0 \alpha}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} {(\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t}\alpha} f {\beta}} g {\gamma})} \gamma)} {F_0 \gamma}\\
      &=
      \homrappXtoYatZ
      {\pendappXtoY{G_1} {F_0 \alpha}}
      {(({\homrappXtoYatZ {\homrappXtoYatZ{\pendappXtoY{(\pendlambdaXdotY \alpha {\homrlambdaXatYdotZ f \beta {\homrlambdaXatYdotZ g \gamma {\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} {(\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t}\alpha} f {\beta}} g {\gamma})} \gamma}}})}\alpha} f \beta} g \gamma}))}
      {F_0 \gamma}\\
      &=
      \homrappXtoYatZ
      {\pendappXtoY{G_1} {F_0 \alpha}}
      {(({\homrappXtoYatZ {\homrappXtoYatZ{\pendappXtoY{(\pendlambdaXdotY \alpha {\homrlambdaXatYdotZ f \beta {\homrlambdaXatYdotZ g \gamma {\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {D_t} {(F_0 \alpha)}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} f \beta)} {(F_0 \beta)}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \beta} g \gamma)} {{F_0} \gamma}}}})}\alpha} f \beta} g \gamma}))}
      {F_0 \gamma}\tag{*}\\
      &=
      \homrappXtoYatZ
      {\pendappXtoY{G_1} {F_0 \alpha}}
      {\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {D_t} {(F_0 \alpha)}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} f \beta)} {(F_0 \beta)}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \beta} g \gamma)} {{F_0} \gamma}}      
      {F_0 \gamma}\\
      &=
      {(({\homrappXtoYatZ {\homrappXtoYatZ{\pendappXtoY{(\pendlambdaXdotY {\alpha'} {\homrlambdaXatYdotZ {f'} {\beta'} {\homrlambdaXatYdotZ {g'} {\gamma'} {{\homrappXtoYatZ {\pendappXtoY {G_1} {\alpha'}} {(\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {D_t}{\alpha'}} {f'} {\beta'}} {g'} {\gamma'})} {\gamma'}}}}})}{F_0\alpha}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} f \beta)} {F_0\beta}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \beta} g \gamma)} {F_0\gamma}}))}\\
      &=
      {(({\homrappXtoYatZ {\homrappXtoYatZ{\pendappXtoY{(\pendlambdaXdotY {\alpha'} {\homrlambdaXatYdotZ {f'} {\beta'} {\homrlambdaXatYdotZ {g'} {\gamma'} {{\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {E_t} {(G_0 {\alpha'})}} {(\homrappXtoYatZ {\pendappXtoY {G_1} {\alpha'}} {f'} {\beta'})} {(G_0 \beta')}} {(\homrappXtoYatZ {\pendappXtoY {G_1} {\beta'}} {g'} {\gamma'})} {{G_0} \gamma'}}}}})}{F_0\alpha}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} f \beta)} {F_0\beta}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \beta} g \gamma)} {F_0\gamma}}))}\\
      &= \homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {E_t} {G_0(F_0\alpha)}}
      {({\homrappXtoYatZ {\pendappXtoY {G_1} {F_0 \alpha}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} f \beta)} {F_0 \beta}})}
      {G_0(F_0\beta)}}
       {({\homrappXtoYatZ {\pendappXtoY {G_1} {F_0 \beta}} {(\homrappXtoYatZ {\pendappXtoY {F_1} \beta} g \gamma)} {F_0 \gamma}})}
       {{F_0} \gamma}
    \end{align*}
  \end{enumerate}
\end{definition}

\begin{definition}[Internal Profunctor]
  A profunctor $R$ between internal categories $\cat C$ and $\cat D$
  consists of
  \begin{enumerate}
  \item A span $R_s : \harr {C_0} {D_0}$.
  \item A ternary ``mixed composition'' action:
    $R_t :\pendallXdotY \alpha {\homrallXYtoZ {\alpha'} {\harrapp {C_1} {\alpha}{\alpha'}} {\homrallXYtoZ {\beta'} {\harrapp {R_s} {\alpha'}{\beta'}} {\homrallXYtoZ {\beta} {\harrapp {D_1} {\beta'} {\beta}} {\harrapp {R_s} {\alpha}{\beta}}}}}$

  \item Satisfying unitality (under a $\pendlambdaXdotY \alpha {\homrlambdaXatYdotZ f \beta {}}$):
    \[
    f = \homrappXtoYatZ {\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {R_t} \alpha} {\pendappXtoY {C_r} \alpha} \alpha}f \beta} {\pendappXtoY {D_r} \beta} \beta
    \]

  \item And associativity (both with a large implicit
    $\pendlambdaXdotY \alpha {\homrlambdaXatYdotZ g \beta \homrlambdaXatYdotZ {g'} {\beta'}
    \homrlambdaXatYdotZ {r} {\beta''}\homrlambdaXatYdotZ {f'} {\alpha''}\homrlambdaXatYdotZ {f} {\alpha'}}$)

    % f : a -> a'
    % f' : a' -> a''
    % r : a'' -> b''
    % g' : b'' -> b'
    % g : b' -> b
    %%
    %
    % f o (f' o r o g') o g
    %% and
    % (f o f') o r o (g' o g)

>>>>

  We show that the type theory is sound for models in \emph{virtual
  equipments}, a kind of double category which is known to model not
  just ordinary categories but enriched, internal and generalized
  multi-categories as well. This means that while proofs and
  constructions in our theory read like typical category-theoretic
  constructions, the proofs automatically apply to these generalized
  categories. Our type theory simplifies arguments over the algebraic
  description of virtual equipments by modeling the algebraic
  operation of restriction of profunctors along functors as instead
  the strict syntactic substitution.
\end{abstract}

@@

At this level of generality, the proofs can be difficult to
understand, as they often involve some combination of the original
category theoretic proof with some additional side-conditions that
everythinng is in fact appropriately step-indexed, continuous etc.
%
One approach to taming this complexity is to prove these general
theorems at a higher level of generality, for instance, proving them
in some abstract 2-category of enriched/internal categories, functors
and transformations.
%
For proofs in generalized category theory specifically, the notion of
a \emph{virtual equipment} has been shown to be occupy an algebraic
sweet-spot of being complex enough to express most basic constructions
(functors, profunctors, natural transformations, adjunctions, kan
extensions, etc) but also general enough to encompass internal,
enriched and ``generalized multi-categories'', which include ordinary
flavors of multi-categories that model intuitionistic and
substructural type theories, but also a great deal more, including
double categories, metric spaces and topological spaces.
%
However, there are two difficulties with this approach.
%
First, the abstract arguments can be difficult to understand compapred
to the original argument for ordinary category theory.
%
Second, the algebraic structure of a virtual equipment is quite
unwieldy to manipulate. The abstraction of a generalized natural
transformation is a kind of ``multi-2-cell'' with many inputs and a
single output. In diagrammatic form these are rendered as

@@

We solve both of these issues by developing an alternative \emph{type
theoretic syntax} for the constructions in a virtual equipment.
%
In this syntax the ``multi-2-cell'' above instead becomes a natural
transformation judgment
\[ \Phi \vdash t : S[F(x);G(y)] \]
where the composable string of input arrows is modeled by an ordered,
linear context $\Phi$ and output $S$ and frames $F,G$ are combined in
the output type.
%
We write natural transformations in a natural deduction style: a
transformation is written as a ``parameterized'' element of the output
profunctor. Furthermore the restriction operators for vertical arrows
into horizontal arrows is naturally modeled by a syntactic
substitution.
%
This interpretation of restriction as substitution both matches the
semantic intuition for the restriction operation as well as greatly
simplifies reasoning about composition of transformations.

@@

\subsection{Applications to Generalized Type Theory}

While we mostly focus on the semantics in category theory, due to the
tight connection between category theory and the semantics of type
theory, there are direct applications to the study of ``abstract type
theory'' as well. We demonstrate how we can model single-variable
simple type theories with products and coproducts within SCT. This is
very much in the spirit of logical frameworks \cite{TODO}, since we
can faithfully model the terms of an object theory inside of our
metatheory. Since we have models of our metatheory in virtual double
categories, we establish sound semantics as well. Put succinctly,
category theory itself is the logical framework for the study of type
theories.

@@

\subsection{Products}

We add in parallel two different notions of product: products of sorts
and products of spans, which have the universal property in their
respective categories.

\begin{figure}
  \begin{mathpar}
    \inferrule{}{\Gamma \vdash 1 : \Sort}\and
    \inferrule*{}{\Gamma\pipe\alpha:C \vdash () : 1}\and
    \inferrule*{\Gamma\pipe\alpha:C \vdash A : 1}{A = ()}
    \inferrule{}{\Gamma\pipe \Xi \vdash 1 \isaSet}\and
    \inferrule{}{\Gamma\pipe \Phi \vdash () :  1}\and
    \inferrule{\Gamma\pipe \Phi \vdash M :  1}{M = ()}\\\\

    \inferrule{\Gamma \vdash C : \Sort \and \Gamma \vdash D : \Sort}
              {\Gamma \vdash C \times D : \Sort}\and

    \inferrule{\forall i \in \{1,2\}.~ \Gamma \pipe \alpha:C \vdash A_i : D_i}{\Gamma \pipe \alpha:C \vdash (A_1,A_2) : D_1 \times D_2}

    \inferrule{\Gamma \pipe \alpha:C \vdash A : D_1 \times D_2}{\Gamma \pipe \alpha:C \vdash \pi_i A : D_i}

    \pi_i(A_1,A_2) = A_i\\
    \inferrule{\Gamma\pipe \alpha:C \vdash A :  D_1\times D_2}{A = (\pi_1 A,\pi_2 A)}\\\\

    \inferrule{\Gamma\pipe\Xi \vdash R \isaSet \and \Gamma\pipe\Xi \vdash S\isaSet}
              {\Gamma\pipe\Xi \vdash R \times S\isaSet}\and

    \inferrule{\forall i \in \{1,2\}.~ \Gamma \pipe\Phi \vdash M_i : R_i}{\Gamma \pipe\Phi \vdash (M_1,M_2) : R_1 \times R_2}

    \inferrule{\Gamma \pipe\Phi\vdash M : R_1 \times R_2}{\Gamma \pipe\Phi \vdash \pi_i M : R_i}

    \pi_i(M_1,M_2) = M_i\\
    \inferrule{\Gamma\pipe\Phi \vdash M :  R_1\times R_2}{M = (\pi_1 M,\pi_2 M)}
  \end{mathpar}
  \caption{Products}
  \label{fig:products}
\end{figure}

One additional connective we might desire is a ``parallel'' product of
sets $R \times' S : \harr {C_1 \times C_2} {D_1 \times D_2}$ where $R
: \harr {C_1} {D_1}$ and $R : \harr {C_2} {D_2}$. We don't need to
define this as a primitive though because it is expressible in terms
of the two kinds of products we have above:

\[
\inferrule
{\Gamma \pipe \alpha : C_i; \beta : D_i \vdash R_i \isaSet}
{\Gamma \pipe \alpha : C_1\times C_2; \beta : D_1 \times D_2
  R_1[\pi_1\alpha/\alpha;\pi_1\beta/\beta] \times R_2[\pi_2\alpha/\alpha;\pi_2\beta/\beta]\isaSet}
\]

The empty products give us access now to several traditional notions
of category theory that we couldn't before define, such as objects and
presheaves of either variance:

$$\textrm{Object}\,C = \varr 1 C$$
$$\textrm{Presheaf}\,C = \harr 1 C$$
$$\textrm{coPresheaf}\,C = \harr C 1$$

However, we find that in practice we don't have much use for these
concepts as most theorems about them easily generalize to more general
theorems about arbitrary functors or profunctors.

@@

\begin{enumerate}
\item A functor $\textrm{wp} : \exists (\alpha:\cat B; \beta:\cat E). \punitinXfromYtoZ {\cat B} \alpha {p\beta}$
\item In the slice category: $p(\textrm{wp}(k)) = \pi_-k$
\item With a universal morphism $\epsilon : \pendallXdotY {k} \punitinXfromYtoZ {} {(\pi_-k,k,\punitrefl{}{\pi_-k})} {k}$
\item In the slice category: $\pi_k \epsilon = \id$ (how to express this nicely in SCT?)
\end{enumerate}

@@
